---
title: "R ToothGrowth Dataset Exploratory Data Analysis and Inferential Study"
output: pdf_document
---

This report investigates the response of the length of teeth in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid) (from ToothGrowth R help).

## Exploratory Data Analysis and Data Summary
The ToothGrowth data set consists of the following variables and types:
``` {r, echo=FALSE}
     library(plyr)
     library(ggplot2)
     data(ToothGrowth)
     str(ToothGrowth)
```
This data set has `r sum(!complete.cases(ToothGrowth))` incomplete cases.  Additionally, each variable's summary statistics are
``` {r, echo=FALSE}
     summary(ToothGrowth)
```
The unique values of Vitamin C dosage are `r unique(ToothGrowth$dose)`, so this variable may be considered a factor variable.

## Hypothesis Tests Around Variation of Mean Tooth Length
Here, we will investigate several hypotheses of the effects of Vitamin C dosage and its devlivery method on mean tooth length:

1. No differnce between delivery by orange juice or absorbic acid overall.
2. No difference based on dosage level, regardless of delivery method.
3. No difference based on delivery method for each dosage.

### Question 1
The summary statistics are as follows:
```{r, echo=FALSE}
     statsQ1 <- ddply(ToothGrowth, .(supp), summarize, mean.len=mean(len),
                      var.len=var(len), num.obs=length(len))
     df.Q1 <- with(statsQ1,num.obs[1]+num.obs[2]-2)
     var.pool.Q1 <- with(statsQ1,
                    ((num.obs[1]-1)*var.len[1]+(num.obs[2]-1)*var.len[2])/df.Q1)
     t.Q1 <- with(statsQ1,
                  (mean.len[1]-mean.len[2])/
                       sqrt(var.pool.Q1*(1/num.obs[1]+1/num.obs[2])))
     alpha <- .05
     qt.Q1 <- qt(1-alpha/2, df = df.Q1)
     statsQ1
```
Using a two-sided t-test with $\alpha =$ `r alpha`, we will reject the null hypothesis of the means being equal if the computed t-value is greater than `r qt.Q1`.  For these data, we get $t =$ `r t.Q1`.  Comparing, we fail to reject the null hypothesis.

### Question 2
The summary statistics are as follows:
```{r, echo=FALSE}
     statsQ2 <- ddply(ToothGrowth, .(dose), summarize, mean.len=mean(len),
                      var.len=var(len), num.obs=length(len))
     df.Q2.12 <- with(statsQ2,num.obs[1]+num.obs[2]-2)
     df.Q2.23 <- with(statsQ2,num.obs[2]+num.obs[3]-2)
     df.Q2.13 <- with(statsQ2,num.obs[1]+num.obs[3]-2)     
     var.pool.Q2.12 <- with(statsQ2,
                    ((num.obs[1]-1)*var.len[1]+(num.obs[2]-1)*var.len[2])/df.Q2.12)
     var.pool.Q2.23 <- with(statsQ2,
                    ((num.obs[2]-1)*var.len[2]+(num.obs[3]-1)*var.len[3])/df.Q2.23)
     var.pool.Q2.13 <- with(statsQ2,
                    ((num.obs[1]-1)*var.len[1]+(num.obs[3]-1)*var.len[3])/df.Q2.13)
     t.Q2.12 <- abs(with(statsQ2,
                  (mean.len[1]-mean.len[2])/
                       sqrt(var.pool.Q2.12*(1/num.obs[1]+1/num.obs[2]))))
     t.Q2.23 <- abs(with(statsQ2,
                  (mean.len[2]-mean.len[3])/
                       sqrt(var.pool.Q2.23*(1/num.obs[2]+1/num.obs[3]))))
     t.Q2.13 <- abs(with(statsQ2,
                  (mean.len[1]-mean.len[3])/
                       sqrt(var.pool.Q2.13*(1/num.obs[1]+1/num.obs[3]))))
     qt.Q2.12 <- qt(1-alpha/2, df = df.Q2.12)
     qt.Q2.23 <- qt(1-alpha/2, df = df.Q2.23)
     qt.Q2.13 <- qt(1-alpha/2, df = df.Q2.13)    
     statsQ2
```
Using the same analysis as for Question 1, but for each of the three mean differences, we get the following two-sided t-scores:
```{r, echo=FALSE}
     results.Q2 <- data.frame(mean.diff = c("1-2", "2-3", "1-3"),
                              abs.t.score = c(t.Q2.12, t.Q2.23, t.Q2.13),
                              t.critical = c(qt.Q2.12, qt.Q2.23, qt.Q2.13))
     results.Q2
```
Snce $|t|>t_{critical}$ for all three scenarios, we reject the null hypotheses.

### Question 3
The summary statistics are as follows:
``` {r, echo=FALSE}
     statsQ3 <- ddply(ToothGrowth, .(supp, dose), summarize, 
                    mean.len=mean(len), var.len=var(len), num.obs=length(len))
     df.Q3.14 <- with(statsQ3,num.obs[1]+num.obs[4]-2)
     df.Q3.25 <- with(statsQ3,num.obs[2]+num.obs[5]-2)
     df.Q3.36 <- with(statsQ3,num.obs[3]+num.obs[6]-2)     
     var.pool.Q3.14 <- with(statsQ3,
                    ((num.obs[1]-1)*var.len[1]+(num.obs[4]-1)*var.len[4])/df.Q3.14)
     var.pool.Q3.25 <- with(statsQ3,
                    ((num.obs[2]-1)*var.len[2]+(num.obs[5]-1)*var.len[5])/df.Q3.25)
     var.pool.Q3.36 <- with(statsQ3,
                    ((num.obs[3]-1)*var.len[3]+(num.obs[6]-1)*var.len[6])/df.Q3.36)
     t.Q3.14 <- abs(with(statsQ3,
                  (mean.len[1]-mean.len[4])/
                       sqrt(var.pool.Q3.14*(1/num.obs[1]+1/num.obs[4]))))
     t.Q3.25 <- abs(with(statsQ3,
                  (mean.len[2]-mean.len[5])/
                       sqrt(var.pool.Q3.25*(1/num.obs[2]+1/num.obs[5]))))
     t.Q3.36 <- abs(with(statsQ3,
                  (mean.len[3]-mean.len[6])/
                       sqrt(var.pool.Q3.36*(1/num.obs[3]+1/num.obs[6]))))
     qt.Q3.14 <- qt(1-alpha/2, df = df.Q3.14)
     qt.Q3.25 <- qt(1-alpha/2, df = df.Q3.25)
     qt.Q3.36 <- qt(1-alpha/2, df = df.Q3.36)    
     statsQ3
```
Again, repeating the same t-tests for the three mean differences, we get the following two-sided t-scores:
```{r, echo=FALSE}
     results.Q3 <- data.frame(dose = c("0.5", "1.0", "2.0"),
                              abs.t.score = c(t.Q3.14, t.Q3.25, t.Q3.36),
                              t.critical = c(qt.Q3.14, qt.Q3.25, qt.Q3.36))
     results.Q3
```
From these results, we fail to reject the null hypothesis only when the dosage is 2.0.

## Conclusion
For the t-tests conducted above, firstly, we see that we don't have enough data to reject the claim that the mean tooth length differs based on delivery method overall.  Secondly, we see that the dosage level does make a difference in mean tooth length.  Finally, we see that the mean tooth length is statistically the same only when the dosage is 2.0 when disregarding the delivery method.
